---
title: "10 Insights from West Virginia School Enrollment Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{10 Insights from West Virginia School Enrollment Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)
```

```{r load-packages}
library(wvschooldata)
library(dplyr)
library(tidyr)
library(ggplot2)

theme_set(theme_minimal(base_size = 14))
```

This vignette explores West Virginia's public school enrollment data, surfacing key trends across the Mountain State's 55 county school districts.

---

## 1. West Virginia educates around 250,000 students

West Virginia's public schools serve roughly a quarter million students across 55 county-based school districts -- one of the simplest administrative structures in the nation.

```{r statewide-data}
enr <- fetch_enr_multi(2014:2025)

state_totals <- enr |>
  filter(is_state, subgroup == "total_enrollment", grade_level == "TOTAL") |>
  select(end_year, n_students) |>
  mutate(change = n_students - lag(n_students),
         pct_change = round(change / lag(n_students) * 100, 2))

state_totals
```

```{r statewide-chart}
ggplot(state_totals, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.2, color = "#002855") +
  geom_point(size = 3, color = "#002855") +
  scale_y_continuous(labels = scales::comma,
                     limits = c(0, NA)) +
  labs(
    title = "West Virginia Public School Enrollment (2014-2025)",
    subtitle = "The Mountain State has seen consistent enrollment decline",
    x = "School Year (ending)",
    y = "Total Enrollment"
  )
```

---

## 2. Kanawha County is the largest district

Kanawha County, home to the state capital Charleston, is West Virginia's largest school district -- though even it would be considered mid-sized in many states.

```{r top-districts-data}
enr_2025 <- fetch_enr(2025)

top_10 <- enr_2025 |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL") |>
  arrange(desc(n_students)) |>
  head(10) |>
  select(district_name, county, n_students)

top_10
```

```{r top-districts-chart}
top_10 |>
  mutate(district_name = forcats::fct_reorder(district_name, n_students)) |>
  ggplot(aes(x = n_students, y = district_name)) +
  geom_col(fill = "#002855") +
  scale_x_continuous(labels = scales::comma) +
  labs(
    title = "West Virginia's 10 Largest School Districts (2025)",
    subtitle = "Kanawha County leads, but no district exceeds 30,000 students",
    x = "Total Enrollment",
    y = NULL
  )
```

---

## 3. Small counties dominate the landscape

West Virginia's county-based system means many very small districts. Several counties have fewer than 1,000 students total.

```{r demographics-data}
# Since WV data lacks demographic breakdowns, we'll analyze district size distribution
size_distribution <- enr_2025 |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL") |>
  mutate(size_category = case_when(
    n_students < 1000 ~ "Under 1,000",
    n_students < 2500 ~ "1,000-2,499",
    n_students < 5000 ~ "2,500-4,999",
    n_students < 10000 ~ "5,000-9,999",
    TRUE ~ "10,000+"
  )) |>
  mutate(size_category = factor(size_category,
                                 levels = c("Under 1,000", "1,000-2,499",
                                           "2,500-4,999", "5,000-9,999", "10,000+"))) |>
  group_by(size_category) |>
  summarize(
    n_districts = n(),
    total_students = sum(n_students, na.rm = TRUE),
    .groups = "drop"
  )

size_distribution
```

```{r demographics-chart}
size_distribution |>
  ggplot(aes(x = size_category, y = n_districts, fill = size_category)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = n_districts), vjust = -0.5) +
  scale_fill_brewer(palette = "Blues") +
  labs(
    title = "West Virginia Districts by Size (2025)",
    subtitle = "Most counties have small student populations",
    x = "District Size (students)",
    y = "Number of Districts"
  )
```

---

## 4. The Eastern Panhandle is growing

While most of West Virginia loses population, the Eastern Panhandle -- Berkeley, Jefferson, and Morgan counties near Washington, D.C. -- has experienced growth from suburban spillover.

```{r regional-data}
# Eastern Panhandle counties (DC suburbs)
panhandle <- c("BERKELEY", "JEFFERSON", "MORGAN")

regional_comparison <- enr |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL") |>
  mutate(region = case_when(
    county %in% panhandle ~ "Eastern Panhandle",
    TRUE ~ "Rest of State"
  )) |>
  group_by(end_year, region) |>
  summarize(n_students = sum(n_students, na.rm = TRUE), .groups = "drop")

regional_comparison |>
  pivot_wider(names_from = region, values_from = n_students)
```

```{r regional-chart}
regional_indexed <- regional_comparison |>
  group_by(region) |>
  mutate(index = n_students / first(n_students) * 100)

ggplot(regional_indexed, aes(x = end_year, y = index, color = region)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  geom_hline(yintercept = 100, linetype = "dashed", alpha = 0.5) +
  scale_color_manual(values = c("Eastern Panhandle" = "#4CAF50",
                                 "Rest of State" = "#F44336")) +
  labs(
    title = "Enrollment Trends: Eastern Panhandle vs. Rest of State",
    subtitle = "Indexed to 2014 = 100",
    x = "School Year",
    y = "Enrollment Index",
    color = NULL
  ) +
  theme(legend.position = "bottom")
```

---

## 5. Coal country continues to decline

The southern coalfield counties -- McDowell, Wyoming, Mingo, Logan, and Boone -- have experienced severe enrollment declines as the coal industry has contracted.

```{r growth-data}
# Coal counties
coal_counties <- c("MCDOWELL", "WYOMING", "MINGO", "LOGAN", "BOONE")

growth_analysis <- enr |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL",
         end_year %in% c(2014, 2025)) |>
  group_by(district_name, county) |>
  filter(n() == 2) |>
  summarize(
    y2014 = n_students[end_year == 2014],
    y2025 = n_students[end_year == 2025],
    change = y2025 - y2014,
    pct_change = round((y2025 / y2014 - 1) * 100, 1),
    .groups = "drop"
  ) |>
  arrange(pct_change)

# Show the most declining districts
declining_10 <- head(growth_analysis, 10)
declining_10
```

```{r growth-chart}
declining_10 |>
  mutate(county = forcats::fct_reorder(county, pct_change)) |>
  ggplot(aes(x = pct_change, y = county)) +
  geom_col(fill = "#B71C1C") +
  geom_text(aes(label = paste0(pct_change, "%")), hjust = 1.1, color = "white") +
  labs(
    title = "Fastest-Declining West Virginia Counties (2014-2025)",
    subtitle = "Coal country leads population loss",
    x = "Percent Change",
    y = NULL
  )
```

---

## 6. McDowell County exemplifies Appalachian decline

McDowell County, once a coal mining powerhouse with over 100,000 residents, now has fewer than 2,500 students -- one of the starkest examples of Appalachian population decline.

```{r mcdowell}
mcdowell_trend <- enr |>
  filter(is_district, county == "MCDOWELL",
         subgroup == "total_enrollment", grade_level == "TOTAL") |>
  select(end_year, n_students) |>
  mutate(pct_of_2014 = round(n_students / first(n_students) * 100, 1))

mcdowell_trend
```

---

## 7. High school enrollment is shrinking faster

Analyzing enrollment by grade level reveals that high school grades are shrinking faster than elementary grades, suggesting accelerating population decline.

```{r grade-level}
grade_trends <- enr |>
  filter(is_state, subgroup == "total_enrollment",
         grade_level %in% c("K", "05", "09", "12")) |>
  select(end_year, grade_level, n_students) |>
  pivot_wider(names_from = grade_level, values_from = n_students)

grade_trends
```

---

## 8. Some counties are holding steady

Despite statewide decline, a few counties have maintained or grown enrollment, primarily those near job centers or in the Eastern Panhandle.

```{r growing}
growing <- growth_analysis |>
  filter(pct_change >= 0) |>
  arrange(desc(pct_change))

growing
```

---
## 9. Kindergarten signals future challenges

Kindergarten enrollment serves as a leading indicator. Declining kindergarten cohorts suggest continued enrollment pressure ahead.

```{r kindergarten}
k_trend <- enr |>
  filter(is_state, subgroup == "total_enrollment",
         grade_level == "K") |>
  select(end_year, n_students) |>
  mutate(change = n_students - lag(n_students))

k_trend
```

---

## 10. 55 districts create administrative challenges

West Virginia's county-based system means even tiny counties maintain full district operations. Several counties have student populations smaller than individual schools in other states.

```{r smallest}
smallest <- enr_2025 |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL") |>
  arrange(n_students) |>
  head(10) |>
  select(district_name, county, n_students)

smallest
```

Counties like Wirt, Calhoun, and Pocahontas each maintain a full school district despite having fewer students than many individual elementary schools elsewhere.

---

## Summary

West Virginia's school enrollment data reveals:
- **Persistent decline**: Statewide enrollment has fallen consistently over the past decade
- **Coal country crisis**: Southern coalfield counties have lost 25-40% of students
- **Eastern Panhandle exception**: DC suburb spillover has brought modest growth
- **Small-district challenges**: Many counties have under 2,000 students
- **Future pressure**: Declining kindergarten enrollment signals more challenges ahead

These patterns reflect broader demographic and economic forces reshaping Appalachia.

---

*Data sourced from West Virginia Department of Education (WVDE) School Finance Data.*
