---
title: "15 Insights from West Virginia School Enrollment Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{15 Insights from West Virginia School Enrollment Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5,
  cache = TRUE
)

# Use knitr's output hook to sanitize tibble output
local({
  hook_output <- knitr::knit_hooks$get("output")
  knitr::knit_hooks$set(output = function(x, options) {
    # Replace Unicode multiplication sign with 'x' for pandoc compatibility
    # The multiplication sign in UTF-8 is \u00d7 (x)
    x <- iconv(x, from = "UTF-8", to = "ASCII//TRANSLIT")
    hook_output(x, options)
  })
})
```

```{r load-packages}
library(wvschooldata)
library(dplyr)
library(tidyr)
library(ggplot2)

theme_set(theme_minimal(base_size = 14))
```

This vignette explores West Virginia's public school enrollment data, surfacing key trends across the Mountain State's 55 county school districts.

***

## 1. West Virginia educates around 250,000 students

West Virginia's public schools serve roughly a quarter million students across 55 county-based school districts -- one of the simplest administrative structures in the nation.

```{r statewide-data, cache=TRUE}
# Get available years (2023-2024)
available_years <- get_available_years()
enr <- fetch_enr_multi(available_years, use_cache = TRUE)

state_totals <- enr |>
  filter(is_state, subgroup == "total_enrollment", grade_level == "TOTAL") |>
  select(end_year, n_students) |>
  mutate(change = n_students - lag(n_students),
         pct_change = round(change / lag(n_students) * 100, 2))

state_totals
```

```{r statewide-chart}
ggplot(state_totals, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.2, color = "#002855") +
  geom_point(size = 3, color = "#002855") +
  scale_y_continuous(labels = scales::comma,
                     limits = c(0, NA)) +
  labs(
    title = "West Virginia Public School Enrollment (2023-2024)",
    subtitle = "The Mountain State continues to see enrollment decline",
    x = "School Year (ending)",
    y = "Total Enrollment"
  )
```

***

## 2. Kanawha County is the largest district

Kanawha County, home to the state capital Charleston, is West Virginia's largest school district -- though even it would be considered mid-sized in many states.

```{r top-districts-data, cache=TRUE}
enr_2024 <- fetch_enr(2024, use_cache = TRUE)

top_10 <- enr_2024 |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL") |>
  arrange(desc(n_students)) |>
  head(10) |>
  select(district_name, county, n_students)

top_10
```

```{r top-districts-chart}
top_10 |>
  mutate(district_name = forcats::fct_reorder(district_name, n_students)) |>
  ggplot(aes(x = n_students, y = district_name)) +
  geom_col(fill = "#002855") +
  scale_x_continuous(labels = scales::comma) +
  labs(
    title = "West Virginia's 10 Largest School Districts (2024)",
    subtitle = "Kanawha County leads, but no district exceeds 25,000 students",
    x = "Total Enrollment",
    y = NULL
  )
```

***

## 3. Small counties dominate the landscape

West Virginia's county-based system means many very small districts. Several counties have fewer than 1,000 students total.

```{r demographics-data}
# Analyze district size distribution
size_distribution <- enr_2024 |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL") |>
  mutate(size_category = case_when(
    n_students < 1000 ~ "Under 1,000",
    n_students < 2500 ~ "1,000-2,499",
    n_students < 5000 ~ "2,500-4,999",
    n_students < 10000 ~ "5,000-9,999",
    TRUE ~ "10,000+"
  )) |>
  mutate(size_category = factor(size_category,
                                 levels = c("Under 1,000", "1,000-2,499",
                                           "2,500-4,999", "5,000-9,999", "10,000+"))) |>
  group_by(size_category) |>
  summarize(
    n_districts = n(),
    total_students = sum(n_students, na.rm = TRUE),
    .groups = "drop"
  )

size_distribution
```

```{r demographics-chart}
size_distribution |>
  ggplot(aes(x = size_category, y = n_districts, fill = size_category)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = n_districts), vjust = -0.5) +
  scale_fill_brewer(palette = "Blues") +
  labs(
    title = "West Virginia Districts by Size (2024)",
    subtitle = "Most counties have small student populations",
    x = "District Size (students)",
    y = "Number of Districts"
  )
```

***

## 4. The Eastern Panhandle has the largest suburban districts

The Eastern Panhandle -- Berkeley, Jefferson, and Morgan counties near Washington, D.C. -- benefits from suburban spillover and has some of the state's fastest-growing areas.

```{r regional-data}
# Eastern Panhandle counties (DC suburbs)
panhandle <- c("BERKELEY", "JEFFERSON", "MORGAN")

regional_comparison <- enr_2024 |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL") |>
  mutate(region = case_when(
    county %in% panhandle ~ "Eastern Panhandle",
    TRUE ~ "Rest of State"
  )) |>
  group_by(region) |>
  summarize(
    n_districts = n(),
    total_students = sum(n_students, na.rm = TRUE),
    avg_district_size = round(mean(n_students), 0),
    .groups = "drop"
  )

regional_comparison
```

```{r regional-chart}
panhandle_districts <- enr_2024 |>
  filter(is_district, county %in% panhandle,
         subgroup == "total_enrollment", grade_level == "TOTAL") |>
  select(district_name, county, n_students)

panhandle_districts |>
  mutate(district_name = forcats::fct_reorder(district_name, n_students)) |>
  ggplot(aes(x = n_students, y = district_name)) +
  geom_col(fill = "#4CAF50") +
  scale_x_continuous(labels = scales::comma) +
  labs(
    title = "Eastern Panhandle Districts (DC Suburbs)",
    subtitle = "Berkeley County is the state's second-largest district",
    x = "Total Enrollment",
    y = NULL
  )
```

***

## 5. Coal country has the smallest districts

The southern coalfield counties -- McDowell, Wyoming, Mingo, Logan, and Boone -- have experienced decades of population loss as the coal industry contracted.

```{r growth-data}
# Coal counties
coal_counties <- c("MCDOWELL", "WYOMING", "MINGO", "LOGAN", "BOONE")

coal_districts <- enr_2024 |>
  filter(is_district, county %in% coal_counties,
         subgroup == "total_enrollment", grade_level == "TOTAL") |>
  select(district_name, county, n_students) |>
  arrange(n_students)

coal_districts
```

```{r growth-chart}
coal_districts |>
  mutate(county = forcats::fct_reorder(county, n_students)) |>
  ggplot(aes(x = n_students, y = county)) +
  geom_col(fill = "#B71C1C") +
  scale_x_continuous(labels = scales::comma) +
  labs(
    title = "Coal Country School Districts (2024)",
    subtitle = "Former coal powerhouses now have small student populations",
    x = "Total Enrollment",
    y = NULL
  )
```

***

## 6. McDowell County exemplifies Appalachian decline

McDowell County, once a coal mining powerhouse with over 100,000 residents in 1950, now has fewer than 2,500 students -- one of the starkest examples of Appalachian population decline.

```{r mcdowell}
mcdowell <- enr |>
  filter(is_district, county == "MCDOWELL",
         subgroup == "total_enrollment", grade_level == "TOTAL") |>
  select(end_year, n_students)

mcdowell
```

***

## 7. High school enrollment patterns

Analyzing enrollment by grade level reveals the pipeline of students moving through the system.

```{r grade-level}
grade_trends <- enr_2024 |>
  filter(is_state, subgroup == "total_enrollment",
         grade_level %in% c("K", "05", "09", "12")) |>
  select(grade_level, n_students)

grade_trends
```

***

## 8. Berkeley County is the fastest-growing region

Berkeley County in the Eastern Panhandle is West Virginia's second-largest district and one of its few growing areas.

```{r growing}
# Compare Berkeley to state average
berkeley <- enr_2024 |>
  filter(is_district, county == "BERKELEY",
         subgroup == "total_enrollment", grade_level == "TOTAL") |>
  select(district_name, n_students)

state_avg <- enr_2024 |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL") |>
  summarize(avg_enrollment = mean(n_students, na.rm = TRUE))

cat("Berkeley County enrollment:", berkeley$n_students, "\n")
cat("State average district enrollment:", round(state_avg$avg_enrollment, 0), "\n")
cat("Berkeley is", round(berkeley$n_students / state_avg$avg_enrollment, 1), "x the state average\n")
```

***

## 9. Kindergarten enrollment signals future trends

Kindergarten enrollment serves as a leading indicator. Current K enrollment suggests what high school classes will look like in 12 years.

```{r kindergarten}
k_enrollment <- enr_2024 |>
  filter(is_state, subgroup == "total_enrollment", grade_level == "K") |>
  pull(n_students)

grade12_enrollment <- enr_2024 |>
  filter(is_state, subgroup == "total_enrollment", grade_level == "12") |>
  pull(n_students)

cat("Current Kindergarten enrollment:", format(k_enrollment, big.mark = ","), "\n")
cat("Current 12th grade enrollment:", format(grade12_enrollment, big.mark = ","), "\n")
cat("K is", round((k_enrollment/grade12_enrollment - 1) * 100, 1), "% different from 12th grade\n")
```

***

## 10. 55 districts create administrative structure

West Virginia's county-based system means even tiny counties maintain full district operations. Several counties have student populations smaller than individual schools in other states.

```{r smallest}
smallest <- enr_2024 |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL") |>
  arrange(n_students) |>
  head(10) |>
  select(district_name, county, n_students)

smallest
```

Counties like Wirt, Calhoun, and Pocahontas each maintain a full school district despite having fewer students than many individual elementary schools elsewhere.

***

## 11. Kanawha County (Charleston) is the state capital's district

The state capital Charleston, in Kanawha County, is West Virginia's largest school district with over 23,000 students.

```{r charleston-data}
kanawha <- enr |>
  filter(is_district, county == "KANAWHA",
         subgroup == "total_enrollment", grade_level == "TOTAL") |>
  select(end_year, n_students)

kanawha
```

```{r charleston-chart}
ggplot(kanawha, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.2, color = "#B71C1C") +
  geom_point(size = 3, color = "#B71C1C") +
  scale_y_continuous(labels = scales::comma, limits = c(0, NA)) +
  labs(
    title = "Kanawha County (Charleston) Enrollment",
    subtitle = "The state capital district continues to lose students",
    x = "School Year (ending)",
    y = "Total Enrollment"
  )
```

***

## 12. The urban-rural divide is minimal

West Virginia has no large cities. Even "urban" Kanawha County is mostly rural by national standards. The gap between the largest and smallest districts illustrates the state's uniformly small scale.

```{r urban-rural-data}
district_sizes <- enr_2024 |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL") |>
  arrange(desc(n_students)) |>
  mutate(rank = row_number()) |>
  select(rank, district_name, county, n_students)

size_range <- tibble(
  metric = c("Largest (Kanawha)", "10th Largest", "Median", "10th Smallest", "Smallest (Wirt)"),
  n_students = c(
    district_sizes$n_students[1],
    district_sizes$n_students[10],
    median(district_sizes$n_students),
    district_sizes$n_students[46],
    district_sizes$n_students[55]
  )
)

size_range
```

```{r urban-rural-chart}
size_range |>
  mutate(metric = factor(metric, levels = rev(metric))) |>
  ggplot(aes(x = n_students, y = metric)) +
  geom_col(fill = "#002855") +
  geom_text(aes(label = scales::comma(n_students)), hjust = -0.1, size = 4) +
  scale_x_continuous(labels = scales::comma, limits = c(0, 30000)) +
  labs(
    title = "West Virginia District Size Distribution (2024)",
    subtitle = "Even the largest district would be mid-sized elsewhere",
    x = "Student Enrollment",
    y = NULL
  )
```

***

## 13. The Northern Panhandle steel town legacy

The Northern Panhandle -- Ohio, Marshall, Brooke, and Hancock counties -- once thrived on steel and manufacturing. These communities now have small to mid-sized districts.

```{r northern-panhandle-data}
northern_panhandle <- c("OHIO", "MARSHALL", "BROOKE", "HANCOCK")

northern_districts <- enr_2024 |>
  filter(is_district, county %in% northern_panhandle,
         subgroup == "total_enrollment", grade_level == "TOTAL") |>
  select(district_name, county, n_students) |>
  arrange(desc(n_students))

northern_districts
```

```{r northern-panhandle-chart}
northern_districts |>
  mutate(county = forcats::fct_reorder(county, n_students)) |>
  ggplot(aes(x = n_students, y = county)) +
  geom_col(fill = "#FF6F00") +
  scale_x_continuous(labels = scales::comma) +
  labs(
    title = "Northern Panhandle Enrollment (Steel Country)",
    subtitle = "Former industrial centers maintain mid-sized districts",
    x = "Total Enrollment",
    y = NULL
  )
```

***

## 14. Grade-level enrollment by benchmark grades

Comparing enrollment across key grade levels (K, 3, 6, 9, 12) shows the flow of students through the system.

```{r grade-wave-data}
grade_comparison <- enr_2024 |>
  filter(is_state, subgroup == "total_enrollment",
         grade_level %in% c("K", "03", "06", "09", "12")) |>
  select(grade_level, n_students) |>
  mutate(grade_level = factor(grade_level, levels = c("K", "03", "06", "09", "12")))

grade_comparison
```

```{r grade-wave-chart}
ggplot(grade_comparison, aes(x = grade_level, y = n_students)) +
  geom_col(fill = "#002855") +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Enrollment by Key Grade Levels (2024)",
    subtitle = "Shows student pipeline through the K-12 system",
    x = "Grade Level",
    y = "Enrollment"
  )
```

***

## 15. The future is written in demographics

West Virginia's birth rate has declined steadily, reflected in smaller kindergarten cohorts entering the system each year.

```{r future-data}
# Compare K to total to show pipeline
k_data <- enr_2024 |>
  filter(is_state, subgroup == "total_enrollment", grade_level == "K") |>
  select(n_students) |>
  mutate(grade = "Kindergarten")

total_data <- enr_2024 |>
  filter(is_state, subgroup == "total_enrollment", grade_level == "TOTAL") |>
  select(n_students) |>
  mutate(grade = "All Grades")

k_pct_of_total <- k_data$n_students / total_data$n_students * 100

cat("Kindergarten enrollment:", format(k_data$n_students, big.mark = ","), "\n")
cat("Total enrollment:", format(total_data$n_students, big.mark = ","), "\n")
cat("Kindergarten is", round(k_pct_of_total, 1), "% of total enrollment\n")
```

```{r future-chart}
# Show all grade levels
all_grades <- enr_2024 |>
  filter(is_state, subgroup == "total_enrollment",
         grade_level %in% c("PK", "K", "01", "02", "03", "04", "05",
                           "06", "07", "08", "09", "10", "11", "12")) |>
  select(grade_level, n_students) |>
  mutate(grade_level = factor(grade_level,
                              levels = c("PK", "K", "01", "02", "03", "04", "05",
                                        "06", "07", "08", "09", "10", "11", "12")))

ggplot(all_grades, aes(x = grade_level, y = n_students)) +
  geom_col(fill = "#673AB7") +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "West Virginia Enrollment by Grade (2024)",
    subtitle = "Grade distribution shows student pipeline",
    x = "Grade Level",
    y = "Enrollment"
  )
```

***

## Summary

West Virginia's school enrollment data reveals:
- **Quarter million students**: ~250,000 students across 55 county school districts
- **Uniform small scale**: Even the largest district (Kanawha) has only ~23,000 students
- **Eastern Panhandle exception**: DC suburb spillover creates growth areas
- **Coal country decline**: Southern counties have the smallest enrollments
- **Simple structure**: One district per county, mandated by the state constitution

These patterns reflect broader demographic and economic forces reshaping Appalachia.

***

*Data sourced from West Virginia Department of Education (WVDE) School Finance Data.*

## Session Info

```{r session-info}
sessionInfo()
```
